---
title: "Athletic Performance."
author: "Project in R by Liz Ángel."
output: html_document
---

```{r setup, include=FALSE}
# We start dowloading libraries. 

library(tidyverse)
library(ggplot2)
library(kableExtra)
library(dplyr)
library(scales)
library(knitr)
```

# 📊 Performance Sport.

## 1. Introduction.

This analysis examines the performance of 5 athletes over 8 weeks of training, evaluating improvements, consistency and trends.

## 2. Load and data prepare.

```{r datos}
# Define our data with the next vectors
athletes <- c('Ana', 'Carlos', 'Maria', 'Javier', 'Laura')
weeks <- c("Week_1", "Week_2", "Week_3", "Week_4", "Week_5", "Week_6", "Week_7", "Week_8")



# Performance matrix (100 points)
performance <- matrix(c(
  85, 78, 82, 79, 88, 90, 92, 95,  # Ana
  72, 75, 80, 82, 85, 83, 88, 90,  # Carlos
  90, 88, 85, 87, 89, 91, 94, 96,  # Maria
  65, 70, 75, 78, 80, 82, 85, 88,  # Javier
  88, 85, 87, 89, 91, 90, 93, 95   # Laura
), nrow = 5, byrow = TRUE)



# Create a DataFrame with our athletes and their points per week.
df_performance <- as.data.frame(performance)
colnames(df_performance) <- weeks
rownames(df_performance) <- athletes



# Show initial table with kable styling; in this case "striped" and "hover".
kable(df_performance, caption = "Performance Data by Athlete") %>%
kable_styling(bootstrap_options = c("striped", "hover"))
```

## 3. Exploratory Analysis.

### 3.1 Descriptive Statistics by athletle.

```{r estadisticas-atletas}
# Calculate statistics by athlete
athlete_stats <- df_performance %>%
  rownames_to_column("Athlete") %>% # Now I have "Athlete" add in our column.
  pivot_longer(
    cols = starts_with("Week_"),    # I want to transform just weeks columns.
    names_to = "Week",              # Now every column "week_" is already in a single column, "Week".
    values_to = "Performance") %>%  # All numeric values are include in a new column, "Performance."
  group_by(Athlete) %>%             # Now our Athletes are in a group with their values and weeks.
  summarise(
    Mean = mean(Performance),
    Minimum = min(Performance),
    Maximum = max(Performance),
    Standard_Deviation = sd(Performance),
    Absolute_Improvement = max(Performance) - min(Performance)
  ) %>%
  arrange(desc(Mean))

# The key is recognize our functions.
# Max() it's maximum score.
# Min() it's the minimum score.
# Mean() it's the mean score.
# sd() it's Standar Deviation


# Once again, I use kable() function to make a better table.
kable(athlete_stats, caption = "Statistics by Athlete") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

### 3.2 Evolution chart Week-to-Week.
 
```{r evolucion-semanal}
# Prepare data for ggplot because I want to make a graphic about our data.
df_long <- df_performance %>%
  rownames_to_column("Athlete") %>%
  pivot_longer(cols = starts_with("Week_"),    # Order again our columns with pivot_longer() function.
               names_to = "Week",
               values_to = "Performance") %>%
  mutate(Week_Num = as.numeric(str_remove(Week, "Week_")))
# This part is our key to make a graphic.
# With mutate() fuction make a new colums with the named "Week_Num".
# Within mutate(), use str_remove() to extract numbers from the "Week" column.
# Using as.numeric() convert those extracted text values into actual numbers.
# The result is a new nureic column "Week_Num" that works perfectly for ggplot and graphs.


# Evolution chart.
ggplot(df_long, aes(x = Week_Num,          # Week numbers(1,2,3...n) 
                    y = Performance,       # Performance Score (It's our matrix)
                    color = Athlete,       # Color by Athlete 
                    group = Athlete)) +    # Conect points by athleta.
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  scale_x_continuous(breaks = 1:8, labels = weeks) +  # Position 1 to 8 called "Week_1, Week_2..".
  labs(
    title = "Performance Evolution by Athlete 8 week training tracking",
    x = "Weeks",
    y = "Performance Score 100 points"
  ) +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "bottom")        # Put our color legend in the bottom side.
```

## 4. Improvement and Consistency Analysis.

### 4.1 Weekly progress.

```{r progreso}
# Calculate week to week differences
df_progress <- df_performance %>%
  apply(1, diff) %>%                   # Apply 1 row and 2 columns to make the diference by week..
  t() %>%                              # And I use transpose matrix "t()" because "apply()" changed rows
                                         # to columns and I need to reverse that change.
  
  as.data.frame()                      # Then, this matrix need to be a DataFrame, for that
                                         #as.data.frame() function.

print(df_progress)                     # Now, we can see evolution by week in the athletes.

colnames(df_progress) <- paste0("Diff_Week", 2:8)
df_progress$Athlete <- athletes
kable(df_progress, caption = "Weekly Performance Differences") %>%
kable_styling(bootstrap_options = c("striped","hover"))
```

### 4.2 Consistency Analysis.

```{r consistencia}
# Calculate coefficient of variation (consistency)
df_consistency <- df_performance %>%
  rownames_to_column("Athlete") %>%
  pivot_longer(cols = starts_with("Week_"),
               names_to = "Week",
               values_to = "Performance") %>%
  group_by(Athlete) %>%
  summarise(
    Mean = mean(Performance),
    Standard_Deviation = sd(Performance),           # Using sd() we get Standar Deviation.
    Coefficient_Variation = sd(Performance) / mean(Performance) * 100,   #This is a key, relative variation.
    .groups = 'drop'
  ) %>%
  arrange(Coefficient_Variation)  #Now, the lower CV (Coeficcient Variaton) will be the first in the DataFrame.

kable(df_consistency, caption = "Consistency Analysis (Lower CV = More Consistent)") %>%
  kable_styling(bootstrap_options = "striped")
```

## 5. Key Findings.

### 5.1 Answers to Analysis questions.

```{r}
# Answer our questions.

# Which athlete shows the most improvement?
most_improved <- athlete_stats %>%
  arrange(desc(Absolute_Improvement)) %>%  # The athlete with the greatest improvement will be fisrt.
  slice(1)

# Which athlete is most consistent?
most_consistent <- df_consistency %>%
  arrange(Coefficient_Variation) %>%       # The lowest CV will be first.  
  slice(1)

# Best overall performance?
best_performance <- athlete_stats %>%
  arrange(desc(Mean)) %>%                # The best mean will be first.
  slice(1)

cat("Main Findings:\n\n")
cat("1. Most improved: ", most_improved$Athlete, "(+", most_improved$Absolute_Improvement, "points)\n")
cat("2. Most consistent: ", most_consistent$Athlete, "(CV:", round(most_consistent$Coefficient_Variation, 1), "%)\n")
cat("3. Best average performance: ", best_performance$Athlete, "(", round(best_performance$Mean, 1), "points)\n")
```

## 6. Conclusions

### 6.1 Conclusions.

1. General positive trend in all athletes.
2. Javier Shows the greatest absolute improvement despite starting with the lowest score.
3. Maria maintains the highest performance consistently.
4. Week 8 Shows the best collective performance in the entire weeks.
